A Procedure to Predict the Subcritical Turbulent Onset Criterion Applied to a Modified Hasegawa-Wakatani Model

### 2014-06-19 HW_non-modal.tex
June 19

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
\usepackage{graphicx}
%\usepackage{amssymb}
%\setlength{\parindent}{0in}

@@

@@

More specifically, we model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, 
(2) linear transient growth deterministically amplifies the turbulent energy (or decrease it in wavenumber ranges where linear damping dominates), 
(3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats). 
Optimally, the timescale for the final step would be the nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic linear time. The procedure ultimately produces a growth rate that can be used to predict turbulent properties such as saturation levels and transport rates 
through mixing length arguments. Furthermore, The sign of the growth rate predicts whether or not turbulence can exist -- a positive growth rate predicting turbulence.

@@

In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 300 \rho_s$.
For the 2D version of the model, we replace $-\xi \gradpar^2$ by $\alpha (= k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
In numerical simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
$n$ and $\phi$ components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.

@@

\beq
\label{B_norm_limit0}
\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right)
\eeq
in which case $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
the normality of a matrix or operator~\cite{trefethen2005}. In practice, however, what matters is not how non-normal $\mathbf{B}_k$ is, but how much transient energy growth is possible in the linear
system. The linear HW system is

\beq
\label{lin_HW}
\diff{u_k}{t} = \mathbf{B}_k u_k
\eeq
where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
its Taylor expansion. The solution is dependent on the initial condition $u(0)$. The energy growth as a function of time is

\beq
\label{Gt}
G_k(t) = \frac{||u(t)||^2}{||u(0)||^2} = \frac{||e^{\mathbf{B}_k t} u(0)||^2}{||u(0)||^2}
\eeq
where the operator $|| \cdot ||$ is the $L_2$-norm. Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such optimal growth curves is given by
$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on the initial condition.
Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
The equality holds when $\mathbf{B}_k$ is normal, meaning the normal case provides a lower bound to the maximal energy growth. 
Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that non-modal growth is possible for small adiabaticity parameter $\alpha$. Furthermore,
the short time transient growth becomes larger as $\alpha$ becomes smaller.

@@

One explanation for this behavior is that non-normal matrices or operators have non-orthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
with it. They are complex and take the form:

@@

Linear non-normality is important for the effect it has on turbulence. It is most often cited as a necessary condition for turbulence in subcritical systems. 
But it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis. 
To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
To define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:

@@

\beq
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
\eeq
where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}, which are quadratic in the fluctuating
quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k,k')$ are triadic in the fluctuating quantities and
account for the energy exchange between fluctuations with different $k$'s. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.

@@

\beqar
\label{steady_state}
\gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0
\eeqar
where the equality to zero clearly holds as long as the turbulence is in a steady or quasi-steady state.
From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0$.

@@

A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the \emph{most stable} eigenmode has zero amplitude.
On the other hand, non-normal systems may grow faster when the more stable eigenmode has finite amplitude than when it has none (see Fig.~\ref{max_en_growth}). 
It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
\label{alpha1e-5_gamma_s_t}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
\label{alpha1_gamma_s_t}
\end{figure}

@@

For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
has started to gain acceptance in the plasma community~\cite{baver2002}, although it has generally been used in normal
or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
The case of highly non-normal turbulence and the associated increased turbulent excitation
hasn't been understood as well. In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
but have not realized them as such.

@@

The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
In fact, for high enough values of $\xi$ ($\xi \ge 10^2$), more energy is injected at $k_z = 0$ than at finite $k_z$ (see Fig.~\ref{gamma_max_vs_xi} below for more details).
We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability, which was discovered by Biaskamp and Zeiler~\cite{biskamp1995} and explored further by Drake et al.~\cite{drake1995}. 
The nonlinear instability works as follows: magnetic-field-aligned ($k_z=0$) convective cells transport density across the equilibrium density gradient, setting up $k_z=0$ density structures. 
These structures are unstable to drift waves, which grow as a secondary instability.
The secondary drift waves, which have finite $k_z$, nonlinearly couple to one another and reinforce the original convective cells (or generate new ones), leading to self-sustainment.

Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one.
Biskamp and Drake understood this and wondered how there could be linear energy injection despite the fact that the linear eigenvectors are stable.
The answer to this came when Camargo et al.~\cite{camargo1998} stated that ``some properties of the drift-wave turbulence
previously attributed to nonlinear terms could also be strongly influenced by the nonnormal character of the linear system.'' Although Camargo et al. focused only on the 2D HW model,
based on this statement, it was likely apparent to them that the 3D HW nonlinear instability was a result of linear non-normality. 
Therefore, when $k_z=0$, no matter the value of $\xi$, the adiabaticity parameter $\alpha = k_z^2 \xi = 0$, meaning the linear matrix operator is highly non-normal and subject to transient growth.
So non-normal effects play a role in the 3D HW system for all $\xi$ due to the presence of $k_z=0$ structures.

@@

In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rates $\gamma_{t,k}$.
We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (modal and transient) processes.
First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a common form -- a state variable multiplied by an inverse time scale:
$\tau_{nl}^{-1} \sim v_E k_\perp$.
This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We thus present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale. More specifically, this model is one in which 1) the turbulence begins as a random state, 
2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.

@@

\beq
\label{gamma_nm_calc}
\gamma_{{\rm{nm}},k} = \frac{1}{\tau_{nl,k}} \int_0^{\tau_{nl,k}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left[ \frac{E_k(\tau_{nl,k})}{E_k(0)}\right].
\eeq
Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and set $E_k(0) = 1$ for normalization, we find,

\beq
\label{gamma_nm}
\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}
\eeq
Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals a characteristic linear time scale at all spatial scales~\cite{schekochihin2012}. 
This is a fair assumption in light of our discussion above, in which we showed that $\gamma_{l,k} = - \gamma_{nl,k}$.

There are a several possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
including inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
After trying a few of these different time scales, we have concluded that the one that works best is the inverse of the linear eigenmode frequency $\omega_{s,k}$
-- the imaginary part of the eigenvalues of the linear matrix operator $\mathbf{B}_k$.
Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use. Because of this, unlike the eigenmode growth rates,
the frequency is robust no matter how nonorthogonal the eigenvectors. We thus arrive at our final definition for the non-modal growth rate:

\beq
\label{gamma_omega_def}
\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \} \right].
\eeq
Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ size
of $\mathbf{B}_k$.

@@

The non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
in which the system is fairly non-normal, $\gamma_{\rm{nm},k}$ predicts $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\rm{nm},k}$ diverge, with $\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\rm{nm},k}$, 
but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$. 
This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
of complete randomization by the nonlinearities breaks down. Apparently, the nonlinearities only partially randomize the turbulence in the weakly non-normal case. 
For the most part, the turbulence at a given $k$ consists primarily of the unstable eigenmode, while the stable eigenmode accounts for only a small amount of energy in the turbulence.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predict the growth rate dominance of the $k_z=0$ mode
despite the $k_z=0$ eigenmode stability.}
\label{gamma_max_vs_kz_with_nm}
\end{figure}

>>>>

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[twocolumn,showkeys,superscriptaddress]{revtex4}
\usepackage{graphicx}
%\usepackage{authblk}
%\usepackage{amssymb}
%\setlength{\parindent}{0in}

@@

\affiliation{Department of Physics and Astronomy, University of California, Los Angeles, California 90095-1547, USA}
\affiliation{Lawrence Livermore National Laboratory, Livermore, California 94550, USA}

@@

More specifically, we model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, 
(2) linear transient growth deterministically amplifies the turbulent energy (or depresses it, depending on the wavenumber), 
(3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats). 
Optimally, the timescale for the final step would be the nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic linear time. The procedure ultimately produces a growth rate that can be used to predict turbulent properties such as saturation levels and transport rates 
through mixing length arguments. Furthermore, a finite positive growth rate (of a certain magnitude) indicates that turbulence can exist given a sufficient amount of noise in the system.
Thus, the growth rate provides a turbulent onset criterion, even for subcritical systems.

The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model -- the 2D and 3D Hasegawa-Wakatani model -- 
that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
In Section~\ref{sec_nm_procedure}, we cover the details of our non-modal predictive procedure and show its success in predicting properties of highly non-normal turbulence. 
Finally, in Section~\ref{sec_subcrit_prediction}, we modify the 3D Hasegawa-Wakatani model to make it subcritical and show how our non-modal technique predicts the critical gradient for the onset of turbulence.

@@

In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 300 \rho_s$.
For the 2D version of the model, we replace $-\xi \gradpar^2$ with a single parameter $\alpha$ (definied as $\alpha = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
In numerical simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
$n$ and $\phi$ components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.

@@

\beq
\label{B_norm_limit0}
\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right).
\eeq
Thus, $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
the normality of a matrix or operator~\cite{trefethen2005}. In practice, however, what matters is not how non-normal $\mathbf{B}_k$ is, but how much transient energy growth is possible in the linear
system. The linear HW system is

\beq
\label{lin_HW}
\diff{u_k}{t} = \mathbf{B}_k u_k
\eeq
where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
its Taylor expansion. The solution is dependent on the initial condition $u(0)$. The energy growth as a function of time is

\beq
\label{Gt}
G_k(t) = \frac{||u(t)||^2}{||u(0)||^2} = \frac{||e^{\mathbf{B}_k t} u(0)||^2}{||u(0)||^2}
\eeq
where the operator $|| \cdot ||$ is the $L_2$-norm. Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such optimal growth curves is given by
$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
The equality holds when $\mathbf{B}_k$ is normal, meaning the normal case provides a lower bound to the maximal energy growth. 
Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that non-modal growth is possible for small adiabaticity parameter $\alpha$. Furthermore,
the short time transient growth becomes larger as $\alpha$ becomes smaller.

@@

Non-normality manifests itself in its eigenvectors -- non-normal matrices or operators have non-orthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
with it. They are complex and take the form:

@@

Linear non-normality is important for the effect it has on turbulence. It is most significantly a necessary condition for turbulence in subcritical systems that have only conservative nonlinearities 
(e.g. advective nonlinearities). However, non-normality can also impact non-subcritical systems, which have unstable linear eigenmodes like the HW system. 
Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis.

To illustrate this in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
In order to define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
Specifically, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:

@@

\beq
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
\eeq
where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}, which are quadratic in the fluctuating
quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
The terms in $T(k,k')$ are triadic in the fluctuating quantities and account for the energy exchange between fluctuations with different $k$'s. 
They are also conservative when summed over all wavenumbers: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$~\cite{camargo1995}.

@@

\beqar
\label{steady_state}
\gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0.
\eeqar
The limit of the last expression vanishes -- as indicated -- for steady or quasi-steady state turbulence only.
From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0$.

@@

A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when only the most unstable eigenmode has finite amplitude.
On the other hand, non-normal systems may grow faster when the more stable eigenmodes have finite amplitude than when they have none (see Fig.~\ref{max_en_growth}). 
It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
\label{alpha1e-5_gamma_s_t}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
\label{alpha1_gamma_s_t}
\end{figure}

@@

For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
has started to gain acceptance in the plasma community~\cite{baver2002}, although it has generally been used in normal
or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
The case of highly non-normal turbulence and the associated increased turbulent excitation hasn't been understood as well. 
In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
but have not realized them as such~\cite{biskamp1995,drake1995,scott2002,scott2005,umansky2009,friedman2012b}.

@@

As Biaskamp and Zeiler~\cite{biskamp1995} discovered, the 3D HW model nonlinear simulation has a fascinating property -- 
the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
In fact, for high enough values of $\xi$ ($\xi \ge 10^2$), more energy is injected at $k_z = 0$ than at finite $k_z$ (see Fig.~\ref{gamma_max_vs_xi} below for more details).
We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability~\cite{biskamp1995,drake1995}. 
The nonlinear instability works as follows: magnetic-field-aligned ($k_z=0$) convective cells transport density across the equilibrium density gradient, setting up $k_z=0$ density structures. 
These structures are unstable to drift waves, which grow as a secondary instability.
The secondary drift waves, which have finite $k_z$, nonlinearly couple to one another and reinforce the original convective cells (or generate new ones), leading to self-sustainment.

Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one. This process occurs at $k_z=0$
for which the adiabaticity parameter iz zero ($\alpha = k_z^2 \xi = 0$) meaning the linear matrix operator is highly non-normal and subject to transient growth.
So non-normal effects play a role in the 3D HW system for all $\xi$ due to the presence of $k_z=0$ structures.

@@

In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rates $\gamma_{t,k}$.
We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (transient and possibly modal) processes.
First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a common form -- a state variable multiplied by an inverse time scale:
$\tau_{nl}^{-1} \sim v_E k_\perp$.
This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We thus present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale. More specifically, this model is one in which 1) the turbulence begins as a random state, 
2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.

@@

\beq
\label{gamma_nm_calc}
\gamma_{{\rm{nm}},k} = \frac{1}{\tau_{nl,k}} \int_0^{\tau_{nl,k}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left[ \frac{E_k(\tau_{nl,k})}{E_k(0)}\right].
\eeq
Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and setting $E_k(0) = 1$ for normalization, we find,

\beq
\label{gamma_nm}
\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}.
\eeq

Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals a characteristic linear time scale at all spatial scales~\cite{schekochihin2012}. 
This is a fair assumption in light of our discussion above, in which we showed that $\gamma_{l,k} = - \gamma_{nl,k}$.

There are a several possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
including the inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
After trying a few of these different time scales, we have concluded that the one that works best is the inverse of the linear eigenmode frequency $\omega_{s,k}$
-- the imaginary part of the eigenvalues of the linear matrix operator $\mathbf{B}_k$.
Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use. Because of this, unlike the eigenmode growth rates,
the frequency is robust no matter how nonorthogonal the eigenvectors. We thus arrive at our final definition for the non-modal growth rate:

\beq
\label{gamma_omega_def}
\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right].
\eeq
Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ dimensionality
of $\mathbf{B}_k$.

@@

The peak non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
in which the system is fairly non-normal, $\gamma_{\rm{nm},k}$ predicts $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\rm{nm},k}$ diverge, with $\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\rm{nm},k}$, 
but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$.

This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
of complete randomization by the nonlinearities breaks down. The nonlinearities only partially randomize the turbulence in the weakly non-normal case such that
the turbulence at a given $k$ always consists primarily of the unstable eigenmode. The stable eigenmode never accounts for a significant portion of the energy in the turbulence.
Makwana et al.~\cite{makwana2011} developed a condition based on the ratio of the stable to unstable eigenmode growth rates that can be used to determine the significance of the stable eigenmode. 
They determined that for $\alpha \ge 10^{-1}$, the stable eigenmode receives very little energy, explaining why our procedure does not work well in the normal limit for the 2D HW model.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predicts the growth rate dominance of the $k_z=0$ mode
despite the $k_z=0$ eigenmode stability.}
\label{gamma_max_vs_kz_with_nm}
\end{figure}

We do the same comparison for the 3D HW model as we did for the 2D HW model. However, there is one additional subtelty when considering the 3D model. In the 3D model, for $k_z=0$, the linear
eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\rm{nm},k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
%We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\rm{nm},k_z=0}$ with $\gamma_{s,k_z=0}$
%since all transients die away in the long time limit. 
Thus, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\rm{nm},k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
by the drift waves, not the convective cells, making $1/\omega_{s,k_z=0.02}$ a good proxy for $\tau_{nl,k_z=0}$.
